---
title: "5. LLM cell type annotation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. LLM cell type annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_beam <- requireNamespace("reticulate", quietly = TRUE) &&
  reticulate::py_module_available("beam.datasets")
knitr::opts_chunk$set(eval = have_beam, collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE,
                      dpi = 140, out.width = "100%", fig.width = 7, fig.height = 4.5)
```

The cell type annotation benchmark of Hou and Ji (2024, Nature Methods, DOI [10.1038/s41592-024-02235-4](https://doi.org/10.1038/s41592-024-02235-4)) scores GPT-4 and GPT-3.5 against three reference-based annotators (CellMarker2.0, SingleR, ScType) by agreement with a manual annotation. The two metrics are the mean agreement score and the full match rate, both higher is better. The R functions call the Python `beam` package through reticulate; `install_beam_python()` installs it.

```{r libs}
library(rbeam)
datasets <- reticulate::import("beam.datasets")
cards <- reticulate::import("beam.cards")
mcda <- reticulate::import("beam.mcda")
beam <- reticulate::import("beam")

g <- datasets$load_gptcelltype()
method_names <- unlist(g$method_names)
metrics <- unlist(g$metric_ids)
```

## Complete cases

The five methods of Hou and Ji, on the datasets where all five were run.

```{r block}
head_methods <- c("GPT-4", "GPT-3.5", "CellMarker2.0", "SingleR", "ScType")
idx <- match(head_methods, method_names)
sub <- g$scores[idx, , , drop = FALSE]
complete <- !apply(is.na(sub[, , 1]), 2, any)
block <- sub[, complete, , drop = FALSE]
block_datasets <- unlist(g$dataset_names)[complete]
cat(sprintf("complete block: %d methods, %d datasets, %d metrics\n",
            dim(block)[1], dim(block)[2], dim(block)[3]))
```

## Ranking

`beam_rank` pools the two metrics across the datasets, [ranks](https://imallona.github.io/beam/docs/explanations/aggregation-methods.html), and runs the sensitivity analysis.

```{r rank}
scores <- beam$Scores(
  values = block,
  tool_names = reticulate::tuple(as.list(head_methods)),
  metric_ids = reticulate::tuple(as.list(metrics)),
  dataset_names = reticulate::tuple(as.list(block_datasets)),
  layout = "long"
)
run <- beam_rank(scores, weights = "equal", method = "saw", seed = 0L)
cat("top-ranked under equal weights / SAW:", run$top_tool, "\n")
ord <- order(run$result$ranks)
for (i in ord) {
  cat(sprintf("  %-14s composite %.3f  rank %d\n",
              head_methods[i], run$result$composite[i], run$result$ranks[i]))
}
```

GPT-4 ranks first.

```{r funky, fig.height = 4.5}
plot(run)
```

## Rank sensitivity

[`beam_rank_sensitivity`](https://imallona.github.io/beam/docs/explanations/rank-sensitivity.html) ranks the methods on each dataset under every weighting and aggregation and splits the rank variance.

```{r rank-sensitivity}
ctx <- mcda$registry_context(as.list(metrics), "saw")
rs <- beam_rank_sensitivity(
  block, ctx$polarity,
  normalization = ctx$normalization, bounds = ctx$bounds,
  baselines = ctx$baselines, targets = ctx$targets,
  tool_names = head_methods, dataset_names = block_datasets
)
cat(sprintf("dataset %.2f, weighting %.2f, aggregation %.2f, interaction %.2f\n",
            rs$dataset_share, rs$weighting_share, rs$aggregation_share, rs$interaction_share))
```

## Dataset concordance

[`dataset_concordance`](https://imallona.github.io/beam/docs/explanations/dataset-concordance-and-discrimination.html) ranks the methods within each dataset and compares every pair of per-dataset orderings with Kendall tau-b.

```{r concordance, fig.height = 4}
conc <- run$dataset_concordance
cat(sprintf("mean agreement across datasets (Kendall tau-b): %.2f\n",
            conc$mean_pairwise_tau))
cat("least typical dataset:",
    conc$dataset_names[[conc$most_idiosyncratic_dataset + 1]], "\n")
beam_plot(run, "dataset_struggle")
```

## Specification curve

[`beam_specification_curve`](https://imallona.github.io/beam/docs/explanations/rank-sensitivity.html#the-specification-curve) lists the rankings of the grid and how often the top method is the same.

```{r specification-curve}
curve <- beam_specification_curve(rs)
top <- curve$tool_names[[curve$most_frequent_top_tool + 1]]
cat(sprintf("%d specifications; %s ranks first in %.0f%%, %d method(s) reach the top\n",
            curve$n_specifications, top, 100 * curve$most_frequent_top_fraction,
            curve$n_distinct_top_tools))
```

```{r specification-curve-plot, fig.height = 5}
beam_plot(curve, "specification_curve")
```

## Blind analysis

[Blinding](https://imallona.github.io/beam/docs/explanations/analysis-blinding.html) fixes the pipeline before the method names are known. `beam_blind` masks the names; `beam_unblind` restores them.

```{r blinding}
blinded <- beam_blind(scores, seed = 0L)
blind_run <- beam_rank(blinded$scores, weights = "equal", method = "saw",
                       sensitivity = FALSE, seed = 0L)
restored <- beam_unblind(blind_run, blinded$seal)
cat("top method after unblinding:", restored$top_tool, "\n")
cat("blinding fingerprint:",
    substr(blind_run$manifest[["blinding"]][["seal_sha256"]], 1, 12), "\n")
```

## Critical difference on agreement

[`beam_critical_difference`](https://imallona.github.io/beam/docs/explanations/critical-difference.html) runs the Friedman test and the Nemenyi post-hoc on the agreement scores of the complete cases.

```{r cd}
agreement <- block[, , 1]
cd <- beam_critical_difference(agreement, higher_is_better = TRUE, tool_names = head_methods)
cat(sprintf("Friedman p = %.1e, Nemenyi CD = %.2f\n",
            cd$friedman_pvalue, cd$critical_difference))
beam_plot(cd, "critical_difference")
```

GPT-4 and CellMarker2.0 are each separate from the other methods; SingleR, ScType and GPT-3.5 are in one clique. See `?beam_rank`, `?beam_rank_sensitivity`, `?beam_critical_difference`.
