---
title: "4. Cross-benchmark meta-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Cross-benchmark meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_beam <- requireNamespace("reticulate", quietly = TRUE) &&
  reticulate::py_module_available("beam.datasets")
have_lme4 <- have_beam && requireNamespace("lme4", quietly = TRUE)
have_netmeta <- have_beam && requireNamespace("netmeta", quietly = TRUE) &&
  requireNamespace("meta", quietly = TRUE)
knitr::opts_chunk$set(eval = have_beam, collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE,
                      dpi = 140, out.width = "100%", fig.width = 7, fig.height = 4.5)
```

Single-cell data-integration benchmarks rank the same methods differently. Five bundled benchmarks are compared here. Four have the scIB metrics (ARI, ASW, kBET, LISI) for five classical methods. BatchBench (Chazarra-Gil et al. 2021, [10.1093/nar/gkab004](https://doi.org/10.1093/nar/gkab004), scores provided by Ruben Chazarra-Gil) has two entropy metrics and is used as within-benchmark ranks. The R functions call the Python `beam` package through reticulate; `install_beam_python()` installs it.

```{r libs}
library(rbeam)
datasets <- reticulate::import("beam.datasets")

canon <- c("combat", "harmony", "fastmnn", "scanorama", "liger")
bench <- c("Tran", "scIB", "OpenProblems")
metric_labels <- c("ARI", "ASW", "kBET", "LISI")
metric_groups <- c("bio", "bio", "batch", "batch")

ib <- datasets$load_integration_benchmarks()
published <- datasets$load_integration_published_ranks()
benchmark <- as.character(unlist(ib$benchmark))
method <- as.character(unlist(ib$method))
metric <- as.character(unlist(ib$metric))
rank_value <- as.numeric(unlist(ib$rank))
```

## Reported rankings

Each benchmark reported a ranking of the five methods with its own metrics, scaling and weighting.

```{r reported}
reported <- sapply(bench, function(b) sapply(canon, function(m) published[[b]][[m]]))
reported
rhos <- combn(bench, 2, function(p) cor(reported[, p[1]], reported[, p[2]], method = "spearman"))
cat(sprintf("mean cross-benchmark Spearman of the reported ranks: %.2f\n", mean(rhos)))
```

## Re-ranking with one rule

The three benchmarks are re-ranked from their raw scores with one rule: four shared metrics, equal weights, ranks within the five methods. The bump plot has the reported rank of each method in the three benchmarks and, right of the divider, the consensus.

```{r bump, fig.height = 4}
mean_rank <- tapply(rank_value, list(benchmark, method), mean)[bench, canon]
consensus <- rank(colMeans(mean_rank), ties.method = "first")
ranks_array <- cbind(reported, beam = consensus[canon])

columns <- c("Tran (reported)", "scIB (reported)", "OpenProblems (reported)", "beam consensus")
beam_rank_bump(
  canon, columns, ranks_array, divider_after = 3,
  title = "Reported method ranks across three benchmarks, and the beam consensus"
)
```

```{r consistent-spearman}
beam_rhos <- combn(bench, 2, function(p) {
  cor(mean_rank[p[1], ], mean_rank[p[2], ], method = "spearman")
})
cat(sprintf("reported: %+.2f   beam consistent: %+.2f\n", mean(rhos), mean(beam_rhos)))
```

## Funky heatmaps per benchmark

The [funky heatmap](https://imallona.github.io/beam/docs/explanations/funky-heatmaps-and-robustness.html) of each benchmark. Each circle is the rank of the method on that metric within the benchmark, coloured by the scIB biological and batch groups.

```{r funky, fig.height = 3.6}
mean_rank_3d <- tapply(rank_value, list(benchmark, method, metric), mean)
for (b in bench) {
  norm <- sapply(metric_labels, function(mk) {
    rr <- mean_rank_3d[b, canon, mk]
    1 - (rr - 1) / (length(canon) - 1)
  })
  composite <- rowMeans(norm, na.rm = TRUE)
  order_ <- rank(-composite, ties.method = "first")
  print(beam_funky_table(
    norm, canon, metric_labels, composite, order_,
    metric_groups = metric_groups,
    title = sprintf("%s: standing of the common methods per metric", b)
  ))
}
```

## BatchBench

BatchBench has two entropy metrics: batch entropy, higher when the batches are better mixed, and cell-type entropy, higher when the cell types are less separated.

```{r tradeoff, fig.height = 4.5}
bb <- datasets$load_batchbench()
bi <- match("batch_entropy", unlist(bb$metric_ids))
ci <- match("cell_type_entropy", unlist(bb$metric_ids))
tradeoff <- data.frame(
  method = unlist(bb$method_names),
  batch = apply(bb$scores[, , bi], 1, mean),
  cell = apply(bb$scores[, , ci], 1, mean)
)
ggplot2::ggplot(tradeoff, ggplot2::aes(batch, cell, label = method)) +
  ggplot2::geom_point(size = 3, colour = "#4477aa") +
  ggplot2::geom_text(hjust = -0.15, size = 3) +
  ggplot2::expand_limits(x = max(tradeoff$batch) + 0.08) +
  ggplot2::labs(x = "batch entropy (higher = batches better mixed)",
                y = "cell-type entropy (higher = cell types more blurred)",
                title = "BatchBench: batch mixing against biology") +
  ggplot2::theme_minimal(base_size = 11)
```

A method low and to the right mixes batches and keeps the cell types separate.

## Network meta-analysis

A [network meta-analysis](https://imallona.github.io/beam/docs/explanations/network-meta-analysis.html) pools the five sources into one ranking. Each method is a treatment and each (benchmark, dataset) pair a study. It needs netmeta.

```{r netmeta, eval = have_netmeta, fig.height = 3.6}
arms <- ib$network_arms()
nma <- beam_network_meta_analysis(arms[[1]], arms[[2]], arms[[3]], arms[[4]], arms[[5]])
beam_plot(nma, "network_forest")
cat(sprintf("I-squared %.2f; inconsistency Q = %.1f on %.0f df, p = %.1e\n",
            nma$i2, nma$q_total, nma$df_total, nma$pval_total))
```

Pooled mean-rank difference of each method from the combat reference, with its 95 percent interval, and its P-score.

## Variance between benchmarks

The method-by-benchmark component of the [variance decomposition](https://imallona.github.io/beam/docs/explanations/attribution-synthesis.html) measures how differently the benchmarks rank the methods. The model is fitted on the four scIB-metric sources, then again with BatchBench added. It needs lme4.

```{r source-variance, eval = have_lme4, message = FALSE, warning = FALSE}
rec <- ib$mean_rank_records()
m <- unlist(rec[[1]]); d <- unlist(rec[[2]]); b <- unlist(rec[[3]]); s <- unlist(rec[[4]])
share <- function(sv) {
  sv$variance_components[["method:benchmark"]] / sum(unlist(sv$variance_components))
}
keep <- b != "BatchBench"
sv4 <- beam_source_variance_decomposition(m[keep], d[keep], b[keep], s[keep])
sv5 <- beam_source_variance_decomposition(m, d, b, s)
cat(sprintf("method-by-benchmark variance share: %.2f without BatchBench, %.2f with it\n",
            share(sv4), share(sv5)))
```

The method-by-benchmark component is 0.23 of the variance with four sources and 0.42 with BatchBench. Both fits are singular: with so few benchmarks at least one variance component is estimated at zero.

## Human pancreas in two pipelines

Tran's Dataset 4 has five pancreas studies and the scIB pancreas task has six; four of the studies are shared. The two pipelines are compared on these overlapping pancreas data.

```{r pancreas, fig.height = 4}
pc <- datasets$load_pancreas_contrast()
pc_methods <- unlist(pc$methods)
tops <- pc$top_method()
cat(sprintf("top method: Tran D4 = %s, scIB pancreas = %s; Spearman of the orders = %+.2f\n",
            tops[[1]], tops[[2]], pc$spearman()))

contrast <- data.frame(
  method = rep(pc_methods, 2),
  pipeline = rep(c("Tran D4", "scIB pancreas"), each = length(pc_methods)),
  rank = c(unlist(pc$tran_mean_rank), unlist(pc$scib_mean_rank))
)
ggplot2::ggplot(contrast, ggplot2::aes(method, rank, fill = pipeline)) +
  ggplot2::geom_col(position = "dodge") +
  ggplot2::scale_fill_manual(values = c("Tran D4" = "#4477aa", "scIB pancreas" = "#ee6677"),
                             name = NULL) +
  ggplot2::labs(x = NULL, y = "mean rank across 4 metrics (lower ranks first)",
                title = "Overlapping pancreas data, two pipelines") +
  ggplot2::theme_minimal(base_size = 11)
```

## Specification curve per benchmark

The [specification curve](https://imallona.github.io/beam/docs/explanations/rank-sensitivity.html#the-specification-curve) lists the rankings of every weighting and aggregation on the mean-rank matrix of each benchmark, and how often the top method is the same.

```{r spec-curve}
mcda <- reticulate::import("beam.mcda")
polarity <- as.list(rep("lower_is_better", length(metric_labels)))
cat("specification curve per benchmark (weighting x aggregation on the mean-rank matrix):\n")
for (b in bench) {
  mm <- ib$method_metric_matrix(b)
  rs <- mcda$rank_sensitivity(mm[[3]], polarity, tool_names = as.list(canon))
  curve <- mcda$specification_curve(rs)
  top <- curve$tool_names[[curve$most_frequent_top_tool + 1]]
  cat(sprintf("  %-12s %-10s first in %3.0f%% of %d combinations, %d method(s) reach the top\n",
              b, top, 100 * curve$most_frequent_top_fraction,
              curve$n_specifications, curve$n_distinct_top_tools))
}
```

## Blind analysis

A [blind analysis](https://imallona.github.io/beam/docs/explanations/analysis-blinding.html) masks the method names before the pipeline is fixed. `beam_blind` shuffles and relabels the rows; `beam_unblind` maps them back.

```{r blind}
beam <- reticulate::import("beam")
mm <- ib$method_metric_matrix("Tran")
scores <- beam$Scores(values = mm[[3]],
                      tool_names = reticulate::tuple(as.list(canon)),
                      metric_ids = reticulate::tuple(as.list(unlist(mm[[2]]))),
                      dataset_names = NULL, layout = "wide")
bl <- beam_blind(scores, seed = 0L)
restored <- beam_unblind(bl$scores, bl$seal)
cat("blinded labels:", paste(as.character(unlist(bl$scores$tool_names)), collapse = ", "), "\n")
cat("restored names:", paste(as.character(unlist(restored$tool_names)), collapse = ", "), "\n")
cat("seal fingerprint:", substr(bl$seal$fingerprint, 1, 12), "\n")
```

## Dataset discrimination

The benchmark of Shen, He and Guan (2026, [10.1371/journal.pcbi.1014008](https://doi.org/10.1371/journal.pcbi.1014008)) has only harmony and scanorama in common with the five methods. [`beam_dataset_discrimination`](https://imallona.github.io/beam/docs/explanations/dataset-concordance-and-discrimination.html#dataset-discrimination) reports, per (dataset, scenario) unit, the separation between the scored methods.

```{r discrimination}
shen <- datasets$load_semisupervised_integration()
shen_methods <- as.character(unlist(shen$method_names))
shen_units <- as.character(unlist(shen$unit_names))
shen_polarity <- as.character(unlist(shen$polarity))

disc <- beam_dataset_discrimination(shen$scores, shen_polarity, dataset_ids = shen_units)
cat(sprintf("%d units, %d methods\n", length(shen_units), length(shen_methods)))
cat(sprintf("mean discrimination (spread): %.3f, mean concordance (W): %.2f\n",
            disc$mean_spread, disc$mean_kendall_w))
cat("most discriminating:", disc$most_discriminating, "\n")
cat("least discriminating:", disc$least_discriminating, "\n")
```

```{r discrimination-plot, fig.height = 5}
beam_plot(disc, "dataset_discrimination", top = 20)
```

Each bar is the spread of the method scores of a unit, coloured by Kendall's W.

`beam_difficulty_concordance` correlates the difficulty of the units for classical and for deep-learning methods.

```{r difficulty}
dl <- c("scVI", "scANVI", "scGEN", "scDREAMER", "scCRAFT", "ItClust")
families <- ifelse(shen_methods %in% dl, "DL", "classical")
conc <- beam_difficulty_concordance(shen$scores, shen_polarity, families,
                                    dataset_ids = shen_units)
cat(sprintf("DL vs classical difficulty concordance: %+.2f\n", conc$concordance[1, 2]))
cat("families disagree most on:", conc$most_divergent_dataset, "\n")
```

```{r difficulty-plot, fig.height = 5}
beam_plot(conc, "difficulty_concordance")
```

Each point is a unit, at the classical score against the deep-learning score.

scIB scores deep-learning methods on the same datasets and metrics as the classical five. `load_scib_integration_families` returns both sets.

```{r scib_families}
fam <- datasets$load_scib_integration_families()
fam_polarity <- as.character(unlist(fam$polarity))
fam_families <- as.character(unlist(fam$families))
scib_conc <- beam_difficulty_concordance(fam$scores, fam_polarity, fam_families,
                                         dataset_ids = as.character(unlist(fam$dataset_names)))
cat(sprintf("scIB DL vs classical concordance (identical data): %+.2f\n",
            scib_conc$concordance[1, 2]))
```

See `?beam_dataset_discrimination`, `?beam_difficulty_concordance`.

