---
title: "4. Cross-benchmark meta-analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Cross-benchmark meta-analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_beam <- requireNamespace("reticulate", quietly = TRUE) &&
  reticulate::py_module_available("beam.datasets")
have_lme4 <- have_beam && requireNamespace("lme4", quietly = TRUE)
have_netmeta <- have_beam && requireNamespace("netmeta", quietly = TRUE) &&
  requireNamespace("meta", quietly = TRUE)
knitr::opts_chunk$set(eval = have_beam, collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE,
                      dpi = 140, out.width = "100%", fig.width = 7, fig.height = 4.5)
```

Single-cell data-integration benchmarks reach different conclusions about which method to use. This vignette asks where the disagreement comes from: the methods, the datasets, or the benchmarker's own analysis choices. beam harmonizes five benchmarks, all bundled with the package. Four of them score on the shared scIB metric family (ARI, ASW, kBET, LISI) for the five classical methods. The fifth, BatchBench (Chazarra-Gil et al. 2021, [10.1093/nar/gkab004](https://doi.org/10.1093/nar/gkab004), with the scores provided by the author), uses its own batch and cell-type entropy and enters on the within-benchmark rank scale. A sixth source, the deep-learning benchmark of Shen, He and Guan (2026, [10.1371/journal.pcbi.1014008](https://doi.org/10.1371/journal.pcbi.1014008)), shares too few methods to rank-harmonize and enters the last section on the method-agnostic discrimination axis instead. The R functions forward to the Python `beam` core; install it once with `install_beam_python()`.

```{r libs}
library(rbeam)
datasets <- reticulate::import("beam.datasets")

canon <- c("combat", "harmony", "fastmnn", "scanorama", "liger")
bench <- c("Tran", "scIB", "OpenProblems")
metric_labels <- c("ARI", "ASW", "kBET", "LISI")
metric_groups <- c("bio", "bio", "batch", "batch")

ib <- datasets$load_integration_benchmarks()
published <- datasets$load_integration_published_ranks()
benchmark <- as.character(unlist(ib$benchmark))
method <- as.character(unlist(ib$method))
metric <- as.character(unlist(ib$metric))
rank_value <- as.numeric(unlist(ib$rank))
```

## The reported rankings disagree

Each benchmark reported its own ranking of the five methods, using its own metrics, scaling and weighting.

```{r reported}
reported <- sapply(bench, function(b) sapply(canon, function(m) published[[b]][[m]]))
reported
rhos <- combn(bench, 2, function(p) cor(reported[, p[1]], reported[, p[2]], method = "spearman"))
cat(sprintf("mean cross-benchmark Spearman of the reported ranks: %.2f\n", mean(rhos)))
```

## beam's consistent re-ranking

beam re-ranks all three from their raw scores with one rule: four shared metrics, equal weight, ranked within the common methods. The bump (subway) plot shows each method's reported rank in the three benchmarks, then the pooled beam consensus on the right of the divider.

```{r bump, fig.height = 4}
mean_rank <- tapply(rank_value, list(benchmark, method), mean)[bench, canon]
consensus <- rank(colMeans(mean_rank), ties.method = "first")
ranks_array <- cbind(reported, beam = consensus[canon])

columns <- c("Tran (reported)", "scIB (reported)", "OpenProblems (reported)", "beam consensus")
beam_rank_bump(
  canon, columns, ranks_array, divider_after = 3,
  title = "Reported method ranks across three benchmarks, and the beam consensus"
)
```

```{r consistent-spearman}
beam_rhos <- combn(bench, 2, function(p) {
  cor(mean_rank[p[1], ], mean_rank[p[2], ], method = "spearman")
})
cat(sprintf("reported: %+.2f   beam consistent: %+.2f\n", mean(rhos), mean(beam_rhos)))
```

Agreement rises from near zero to moderate. Much of the disagreement was in the analysis choices, not the methods.

## Funky heatmaps per benchmark

The glyph tables show why the reported orders differ: the per-metric pattern over the five methods is not the same across benchmarks. Each circle is the method's standing on that metric within the benchmark (larger is better), coloured by the scIB biological and batch groups.

```{r funky, fig.height = 3.6}
mean_rank_3d <- tapply(rank_value, list(benchmark, method, metric), mean)
for (b in bench) {
  norm <- sapply(metric_labels, function(mk) {
    rr <- mean_rank_3d[b, canon, mk]
    1 - (rr - 1) / (length(canon) - 1)
  })
  composite <- rowMeans(norm, na.rm = TRUE)
  order_ <- rank(-composite, ties.method = "first")
  print(beam_funky_table(
    norm, canon, metric_labels, composite, order_,
    metric_groups = metric_groups,
    title = sprintf("%s: standing of the common methods per metric", b)
  ))
}
```

## A fifth source with a different lens: BatchBench

BatchBench scores the methods with two entropy metrics instead of the scIB family: batch entropy, where a higher value means the batches are better mixed, and cell-type entropy, where a higher value means the cell types are more blurred, so a lower value is better for biology. Ruben Chazarra-Gil provided the scores by personal communication. The two metrics show the batch-mixing against biology tradeoff directly.

```{r tradeoff, fig.height = 4.5}
bb <- datasets$load_batchbench()
bi <- match("batch_entropy", unlist(bb$metric_ids))
ci <- match("cell_type_entropy", unlist(bb$metric_ids))
tradeoff <- data.frame(
  method = unlist(bb$method_names),
  batch = apply(bb$scores[, , bi], 1, mean),
  cell = apply(bb$scores[, , ci], 1, mean)
)
ggplot2::ggplot(tradeoff, ggplot2::aes(batch, cell, label = method)) +
  ggplot2::geom_point(size = 3, colour = "#4477aa") +
  ggplot2::geom_text(hjust = -0.15, size = 3) +
  ggplot2::expand_limits(x = max(tradeoff$batch) + 0.08) +
  ggplot2::labs(x = "batch entropy (higher = batches better mixed)",
                y = "cell-type entropy (higher = cell types more blurred)",
                title = "BatchBench: batch mixing against biology") +
  ggplot2::theme_minimal(base_size = 11)
```

A method low and to the right mixes batches well while keeping the cell types separate. harmony sits high and to the right: it mixes batches well but blurs the cell types more than the others. A benchmark that weights biology equally then ranks it below fastMNN, which is what BatchBench does.

## One coherent ranking: network meta-analysis

A network meta-analysis pools all five sources into one ranking, the way clinical research ranks treatments never all tried in one trial. Each method is a treatment, each (benchmark, dataset) block is a study. It needs R's netmeta.

```{r netmeta, eval = have_netmeta, fig.height = 3.6}
arms <- ib$network_arms()
nma <- beam_network_meta_analysis(arms[[1]], arms[[2]], arms[[3]], arms[[4]], arms[[5]])
beam_plot(nma, "network_forest")
cat(sprintf("I-squared %.2f; inconsistency Q = %.1f on %.0f df, p = %.1e\n",
            nma$i2, nma$q_total, nma$df_total, nma$pval_total))
```

Each method's pooled mean-rank difference from the combat reference is shown with its 95 percent interval; an interval left of the zero line outranks the reference. The P-score is the probability the method outranks a random competitor.

With BatchBench in the pool, harmony no longer leads on its own: it ties with liger at the top, fastMNN follows, and combat trails. The high I-squared and the significant inconsistency Q say the direct and the indirect evidence do not line up, the network reading of the residual disagreement.

## How much of the variance is the benchmark

A cross-source variance decomposition puts a number on the split: the method-by-benchmark component is the disagreement attributable to the benchmark rather than the method. The chunk fits it with the four scIB-family sources and again with BatchBench added, so the effect of the fifth source is visible. It needs R's lme4.

```{r source-variance, eval = have_lme4, message = FALSE, warning = FALSE}
rec <- ib$mean_rank_records()
m <- unlist(rec[[1]]); d <- unlist(rec[[2]]); b <- unlist(rec[[3]]); s <- unlist(rec[[4]])
share <- function(sv) {
  sv$variance_components[["method:benchmark"]] / sum(unlist(sv$variance_components))
}
keep <- b != "BatchBench"
sv4 <- beam_source_variance_decomposition(m[keep], d[keep], b[keep], s[keep])
sv5 <- beam_source_variance_decomposition(m, d, b, s)
cat(sprintf("method-by-benchmark variance share: %.2f without BatchBench, %.2f with it\n",
            share(sv4), share(sv5)))
```

Adding BatchBench raises the method-by-benchmark share. Its two entropy metrics weight biological conservation and batch mixing equally, and that stance disagrees with the scIB-family benchmarks about harmony, so more of the spread now sits in the benchmark rather than the method. Part of the rise also comes from BatchBench ranking four methods rather than five, the same scale difference Tyler has, so read it as a clear direction more than an exact number. The fits are singular (with so few benchmarks one or more variance components are estimated at zero), which lme4 reports but which does not stop the share from being computed; it is one more reason to read the value as a direction.

## Same data, two pipelines: the human pancreas

The pooled decomposition mixes pipeline and dataset differences. The one confirmed dataset overlap removes the data confound: Tran's Dataset 4 and scIB's pancreas task are built from the same five studies, so running both pipelines on this shared block reads off the pipeline effect with no data variability in the way.

```{r pancreas, fig.height = 4}
pc <- datasets$load_pancreas_contrast()
pc_methods <- unlist(pc$methods)
tops <- pc$top_method()
cat(sprintf("top method: Tran D4 = %s, scIB pancreas = %s; Spearman of the orders = %+.2f\n",
            tops[[1]], tops[[2]], pc$spearman()))

contrast <- data.frame(
  method = rep(pc_methods, 2),
  pipeline = rep(c("Tran D4", "scIB pancreas"), each = length(pc_methods)),
  rank = c(unlist(pc$tran_mean_rank), unlist(pc$scib_mean_rank))
)
ggplot2::ggplot(contrast, ggplot2::aes(method, rank, fill = pipeline)) +
  ggplot2::geom_col(position = "dodge") +
  ggplot2::scale_fill_manual(values = c("Tran D4" = "#4477aa", "scIB pancreas" = "#ee6677"),
                             name = NULL) +
  ggplot2::labs(x = NULL, y = "mean rank across 4 metrics (lower ranks first)",
                title = "Same pancreas data, two pipelines") +
  ggplot2::theme_minimal(base_size = 11)
```

Both pipelines agree on the leader on the shared data, but the rest of the order disagrees more than the pooled Spearman would predict. The disagreement is not the data; it is in the pipeline.

## Specification curve per benchmark

The specification curve lists the rankings every weighting-by-aggregation combination gives on each benchmark's mean-rank matrix, and reports how often the top method holds. On Tran and scIB the top method holds across all combinations; on OpenProblems it does not.

```{r spec-curve}
mcda <- reticulate::import("beam.mcda")
polarity <- as.list(rep("lower_is_better", length(metric_labels)))
cat("specification curve per benchmark (weighting x aggregation on the mean-rank matrix):\n")
for (b in bench) {
  mm <- ib$method_metric_matrix(b)
  rs <- mcda$rank_sensitivity(mm[[3]], polarity, tool_names = as.list(canon))
  curve <- mcda$specification_curve(rs)
  top <- curve$tool_names[[curve$most_frequent_top_tool + 1]]
  cat(sprintf("  %-12s %-10s first in %3.0f%% of %d combinations, %d method(s) reach the top\n",
              b, top, 100 * curve$most_frequent_top_fraction,
              curve$n_specifications, curve$n_distinct_top_tools))
}
```

## Blind analysis

A blind analysis hides the method names before the pipeline is fixed. These benchmarks rank on the mean-rank scale rather than through metric cards, so the demonstration relabels the matrix and then restores the names with the seal. `beam_blind` shuffles and relabels the rows; `beam_unblind` maps them back.

```{r blind}
beam <- reticulate::import("beam")
mm <- ib$method_metric_matrix("Tran")
scores <- beam$Scores(values = mm[[3]],
                      tool_names = reticulate::tuple(as.list(canon)),
                      metric_ids = reticulate::tuple(as.list(unlist(mm[[2]]))),
                      dataset_names = NULL, layout = "wide")
bl <- beam_blind(scores, seed = 0L)
restored <- beam_unblind(bl$scores, bl$seal)
cat("blinded labels:", paste(as.character(unlist(bl$scores$tool_names)), collapse = ", "), "\n")
cat("restored names:", paste(as.character(unlist(restored$tool_names)), collapse = ", "), "\n")
cat("seal fingerprint:", substr(bl$seal$fingerprint, 1, 12), "\n")
```

## A sixth source with disjoint methods: dataset discrimination

Every analysis above compares method orders, so it needs shared methods. The deep-learning benchmark of Shen, He and Guan (2026, [10.1371/journal.pcbi.1014008](https://doi.org/10.1371/journal.pcbi.1014008)) overlaps the classical five only on harmony and scanorama, too few to rank-harmonize. `beam_dataset_discrimination` reports a property defined per dataset: how much each dataset separates the methods it scores, as the spread of the pooled scores (the effect size) and Kendall's W over the method-by-metric matrix (the consistency). beam treats each (dataset, scenario) pair as one unit.

```{r discrimination}
shen <- datasets$load_semisupervised_integration()
shen_methods <- as.character(unlist(shen$method_names))
shen_units <- as.character(unlist(shen$unit_names))
shen_polarity <- as.character(unlist(shen$polarity))

disc <- beam_dataset_discrimination(shen$scores, shen_polarity, dataset_ids = shen_units)
cat(sprintf("%d units, %d methods\n", length(shen_units), length(shen_methods)))
cat(sprintf("mean discrimination (spread): %.3f, mean concordance (W): %.2f\n",
            disc$mean_spread, disc$mean_kendall_w))
cat("most discriminating:", disc$most_discriminating, "\n")
cat("least discriminating:", disc$least_discriminating, "\n")
```

```{r discrimination-plot, fig.height = 5}
beam_plot(disc, "dataset_discrimination", top = 20)
```

Each bar is one unit's spread of method scores, coloured by Kendall's W, the agreement of the metrics on the method order within that unit.

`beam_difficulty_concordance` then splits the methods into two categories, classical and deep learning, and asks whether they find the same datasets hard. A low concordance means the hardness is specific to a kind of method rather than a property of the data.

```{r difficulty}
dl <- c("scVI", "scANVI", "scGEN", "scDREAMER", "scCRAFT", "ItClust")
families <- ifelse(shen_methods %in% dl, "DL", "classical")
conc <- beam_difficulty_concordance(shen$scores, shen_polarity, families,
                                    dataset_ids = shen_units)
cat(sprintf("DL vs classical difficulty concordance: %+.2f\n", conc$concordance[1, 2]))
cat("families disagree most on:", conc$most_divergent_dataset, "\n")
```

```{r difficulty-plot, fig.height = 5}
beam_plot(conc, "difficulty_concordance")
```

Each point is one unit, at the classical score against the deep-learning score. Points off the equality line are hard for one family only.

The Shen concordance is low: degraded labels affect only the methods that use them, so the hard units are harder for the deep-learning methods. On a benchmark whose hardness is dataset complexity the two categories agree. A single cross-method ranking would hide which kind a benchmark measures.

The scIB source harmonized above scores deep-learning methods on the same datasets and metrics as the classical five, so the only thing that changes between classical and deep learning is the method. `load_scib_integration_families` returns both blocks, a controlled version of the same contrast.

```{r scib_families}
fam <- datasets$load_scib_integration_families()
fam_polarity <- as.character(unlist(fam$polarity))
fam_families <- as.character(unlist(fam$families))
scib_conc <- beam_difficulty_concordance(fam$scores, fam_polarity, fam_families,
                                         dataset_ids = as.character(unlist(fam$dataset_names)))
cat(sprintf("scIB DL vs classical concordance (identical data): %+.2f\n",
            scib_conc$concordance[1, 2]))
```

On scIB the two categories agree (about 0.90), as on OpenProblems and unlike Shen, with the datasets and metrics held fixed, so the low Shen concordance is a property of annotation degradation rather than of being a deep-learning method. See `?beam_dataset_discrimination`, `?beam_difficulty_concordance`.

## Reading

This vignette measures agreement between benchmarks. The single-benchmark vignettes measure agreement between the datasets inside one benchmark, with `beam_dataset_concordance`. A method ordered consistently across the datasets of one benchmark yet ranked differently across benchmarks points at the analysis pipeline rather than the data; the network meta-analysis and source-variance decomposition here are the across-benchmark counterparts of that within-benchmark concordance.

Standardizing the decision making with beam removes part of the apparent disagreement between these benchmarks, but it does not close it: the network meta-analysis finds a significant residual inconsistency even under the shared rule. BatchBench makes the point sharper. It scores biology and batch mixing with equal weight, disagrees with the scIB-family benchmarks about harmony, and so raises the share of the spread that sits in the benchmarker's choices rather than the methods. Harmonizing the analysis narrows the gap rather than erasing it. See `?beam_network_meta_analysis`, `?beam_source_variance_decomposition`.
