---
title: "beam on the Duo 2018 single-cell clustering benchmark"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{beam on the Duo 2018 single-cell clustering benchmark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_beam <- requireNamespace("reticulate", quietly = TRUE) &&
  reticulate::py_module_available("beam.datasets")
knitr::opts_chunk$set(
  eval = have_beam,
  collapse = TRUE,
  comment = "#>"
)
```

This vignette runs the canonical beam pipeline on the Duo et al. 2018 single-cell RNA-seq clustering benchmark (F1000Research, 14 methods x 12 data sets x 4 metrics) bundled with the package.

The R wrappers forward to the Python implementation via reticulate. Install the Python side once with `install_beam_python()` if you have not already.

```{r install, eval = FALSE}
library(rbeam)
install_beam_python()
```

## Load the benchmark

```{r load}
library(rbeam)
py <- reticulate::import("beam.datasets")
duo <- py$load_duo2018()
length(duo$method_names)
length(duo$dataset_names)
duo$metric_ids
```

## Rank the methods

`beam_rank` resolves polarity, normalization, bounds and baselines from the metric cards, runs the MCDA pipeline (default SAW with equal weights), and runs the default sensitivity primitives on the same context.

```{r rank}
scen <- reticulate::import("beam.scenarios")
# Ranking the full Duo tensor needs the dataset-axis construction shown in the
# Python examples/duo2018.qmd. This short vignette demonstrates the wrapper on a
# small bundled scenario; the scores are a bare matrix, so pass the metric ids
# and tool names alongside it.
sc <- scen$dominant_method_scenario()
result <- beam_rank(
  sc$scores,
  weights = "equal",
  method = "saw",
  sensitivity = FALSE,
  metric_ids = sc$metric_ids,
  tool_names = sc$method_names
)
result$top_tool
```

## Render a self-contained HTML report

```{r report, eval = FALSE}
beam_report(result, "duo2018_report.html")
```

The output is a single HTML file with the ranking, the normalization diagnostics, the sensitivity outputs, a critical-difference section when the input is a tensor with more than one dataset, and a plain-language recommendation. Figures are embedded as base64 PNGs; the file has no external dependencies and renders offline.

## Where to go next

- `vignette("duo2018", package = "rbeam")` is intentionally short and CRAN-safe. The full Python vignettes (M4 forecasting, OpenProblems batch integration, the cross-benchmark meta-analysis, the transportation example with partial coverage) live in the beam Python repo under `examples/` and are rendered on the documentation site.
- `?beam_rank` and `?beam_report` for the wrapper docs.
- `?beam_metric_show` to inspect any metric card from the registry.
