---
title: "3. Method-by-dataset heterogeneity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Method-by-dataset heterogeneity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_beam <- requireNamespace("reticulate", quietly = TRUE) &&
  reticulate::py_module_available("beam.datasets")
have_lme4 <- have_beam && requireNamespace("lme4", quietly = TRUE)
have_psychotree <- have_beam && requireNamespace("psychotree", quietly = TRUE)
have_pl <- have_beam && requireNamespace("PlackettLuce", quietly = TRUE)
knitr::opts_chunk$set(eval = have_beam, collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE,
                      dpi = 140, out.width = "100%", fig.width = 7, fig.height = 4.5)
```

[Three models](https://imallona.github.io/beam/docs/explanations/method-by-dataset-heterogeneity.html) of the method-by-dataset interaction, all fitted in R. Each section needs its package (`lme4`, `psychotree`, `PlackettLuce`) and runs only when it is installed. The scores come from the Python `beam` package through reticulate; `install_beam_python()` installs it. The data are the ARI scores of the Duo 2018 benchmark.

```{r libs}
library(rbeam)
datasets <- reticulate::import("beam.datasets")

duo <- datasets$load_duo2018()
method_names <- unlist(duo$method_names)
dataset_names <- unlist(duo$dataset_names)
ari <- duo$tensor(list("ari"))[, , 1]
```

## Mixed-effects

[`beam_mixed_effects`](https://imallona.github.io/beam/docs/explanations/method-by-dataset-heterogeneity.html) fits `score ~ method + (1 | dataset)` in lme4. The dataset intraclass correlation is the fraction of the variance attributable to the dataset intercept.

```{r mixed-effects, eval = have_lme4, fig.height = 4.5}
me <- beam_mixed_effects(ari, method_names, dataset_names)
cat(sprintf("dataset shift (ICC): %.2f of the ARI variance\n", me$icc_dataset))
beam_plot(me, "model_effects", xlabel = "ARI marginal mean (higher ranks first)",
          title = "Mixed-effects method effects on ARI")
```

```{r variance-components, eval = have_lme4, fig.height = 3.4}
beam_plot(me, "variance_components", title = "ARI variance by component")
```

## Dataset concordance on ARI

[`beam_dataset_concordance`](https://imallona.github.io/beam/docs/explanations/dataset-concordance-and-discrimination.html) ranks the methods within each dataset on ARI and compares every pair of per-dataset orderings with Kendall tau-b.

```{r concordance, fig.height = 5}
conc <- beam_dataset_concordance(
  duo$tensor(list("ari")), "higher_is_better",
  dataset_names = dataset_names, tool_names = method_names,
  metric_ids = list("ari"), missing = "worst"
)
cat(sprintf("mean agreement across datasets on ARI (Kendall tau-b): %.2f\n",
            conc$mean_pairwise_tau))
cat("least typical dataset:",
    conc$dataset_names[[conc$most_idiosyncratic_dataset + 1]], "\n")
beam_plot(conc, "dataset_concordance")
```

## Bradley-Terry tree

The [Bradley-Terry tree](https://imallona.github.io/beam/docs/explanations/method-by-dataset-heterogeneity.html#bradley-terry-trees) compares the methods pairwise on each dataset, and `psychotree::bttree` splits the datasets by their features, with a ranking per leaf.

```{r bt-fit, eval = have_psychotree, include = FALSE}
# Fit in a hidden chunk: the per-leaf standard-error step uses a sandwich vcov
# that can write a harmless message on a degenerate leaf. The worths and the
# split decision are unaffected.
feats <- datasets$load_duo2018_features()
aligned <- feats$aligned_to(reticulate::tuple(as.list(dataset_names)))
features <- c(lapply(aligned[[1]], as.numeric), lapply(aligned[[2]], as.character))
bt <- beam_bradley_terry_tree(ari, method_names, dataset_names,
                              features = features, minsize = 4L)
```

```{r bt-show, eval = have_psychotree, fig.height = 3.4}
cat("a dataset feature splits the ranking:", bt$did_split, "\n")
beam_plot(bt, "bradley_terry_leaves", title = "Bradley-Terry tree leaves on ARI")
```

Each bar is a leaf, sized by its number of datasets and labelled by its top method. A tree without a split has one bar.

## Plackett-Luce

[`beam_plackett_luce`](https://imallona.github.io/beam/docs/explanations/method-by-dataset-heterogeneity.html#plackett-luce-full-rankings) fits a Plackett-Luce model to the full ranking of the methods on each dataset. The worths sum to one.

```{r pl, eval = have_pl, fig.height = 4}
pl <- beam_plackett_luce(ari, method_names)
worth <- data.frame(method = method_names, worth = pl$worth)
ggplot2::ggplot(worth, ggplot2::aes(stats::reorder(method, worth), worth)) +
  ggplot2::geom_col(fill = "#228833") +
  ggplot2::coord_flip() +
  ggplot2::labs(x = NULL, y = "Plackett-Luce worth", title = "Plackett-Luce worths on ARI") +
  ggplot2::theme_minimal(base_size = 11)
```

## Cell type annotation

The cell type annotation benchmark of Hou and Ji (2024) has 36 datasets where GPT-4, GPT-3.5 and three reference-based annotators (CellMarker2.0, SingleR, ScType) were all run. Each dataset is a (source, tissue) pair with a species, a normal-or-cancer label and a cell-type count. The metric is the mean agreement with a manual annotation, higher is better.

```{r gpt-block}
g <- datasets$load_gptcelltype()
gpt_methods <- c("GPT-4", "GPT-3.5", "CellMarker2.0", "SingleR", "ScType")
idx <- match(gpt_methods, unlist(g$method_names))
sub <- g$scores[idx, , , drop = FALSE]
complete <- !apply(is.na(sub[, , 1]), 2, any)
agreement <- sub[, complete, 1]
gpt_datasets <- unlist(g$dataset_names)[complete]
cat(sprintf("%d methods, %d datasets\n", nrow(agreement), ncol(agreement)))
```

```{r gpt-mixed, eval = have_lme4, message = FALSE, warning = FALSE}
me <- beam_mixed_effects(agreement, gpt_methods, gpt_datasets)
cat(sprintf("dataset shift (ICC): %.2f of the agreement variance\n", me$icc_dataset))
```

```{r gpt-bt, eval = have_psychotree, fig.height = 4}
gf <- g$features$aligned_to(reticulate::tuple(as.list(gpt_datasets)))
features <- c(lapply(gf[[1]], as.numeric), lapply(gf[[2]], as.character))
bt <- beam_bradley_terry_tree(agreement, gpt_methods, gpt_datasets,
                              features = features, minsize = 6L)
cat("a dataset feature splits the ranking:", bt$did_split, "\n")
beam_plot(bt, "bradley_terry_leaves",
          title = "Bradley-Terry tree leaves on annotation agreement")
```

## Specification curve and blind analysis

The [specification curve](https://imallona.github.io/beam/docs/explanations/rank-sensitivity.html#the-specification-curve) on the Duo ARI, runtime and Shannon entropy difference, over every weighting, aggregation and dataset:

```{r spec-curve}
mcda <- reticulate::import("beam.mcda")
metrics3 <- c("ari", "runtime", "shannon_entropy_diff")
ctx <- mcda$registry_context(as.list(metrics3), "saw")
rs <- beam_rank_sensitivity(duo$tensor(metrics3), ctx$polarity,
                            normalization = ctx$normalization, bounds = ctx$bounds,
                            baselines = ctx$baselines, targets = ctx$targets,
                            missing = "worst", tool_names = method_names,
                            dataset_names = dataset_names)
curve <- beam_specification_curve(rs)
top <- curve$tool_names[[curve$most_frequent_top_tool + 1]]
cat(sprintf("specification curve: %s ranks first in %.0f%% of %d combinations, %d reach the top\n",
            top, 100 * curve$most_frequent_top_fraction,
            curve$n_specifications, curve$n_distinct_top_tools))
```

```{r spec-curve-plot, fig.height = 5}
beam_plot(curve, "specification_curve")
```

[Blinding](https://imallona.github.io/beam/docs/explanations/analysis-blinding.html) fixes the analysis before the method names are known. `beam_blind` hides the names; `beam_unblind` restores them.

```{r blind}
beam <- reticulate::import("beam")
scores <- beam$Scores(values = duo$tensor(metrics3),
                      tool_names = reticulate::tuple(as.list(method_names)),
                      metric_ids = reticulate::tuple(as.list(metrics3)),
                      dataset_names = reticulate::tuple(as.list(dataset_names)),
                      layout = "long")
bl <- beam_blind(scores, seed = 0L)
run <- beam_rank(bl$scores, weights = "equal", method = "saw",
                 missing = "worst", sensitivity = FALSE, seed = 0L)
restored <- beam_unblind(run, bl$seal)
cat("top method after unblinding:", restored$top_tool, "\n")
cat("blinding fingerprint:",
    substr(run$manifest[["blinding"]][["seal_sha256"]], 1, 12), "\n")
```

See `?beam_mixed_effects`, `?beam_bradley_terry_tree`, `?beam_plackett_luce`,
`?beam_specification_curve`, `?beam_blind`.
