---
title: "2. Metric-set diagnostics: validity, reliability, dimensionality"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Metric-set diagnostics: validity, reliability, dimensionality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_beam <- requireNamespace("reticulate", quietly = TRUE) &&
  reticulate::py_module_available("beam.datasets")
knitr::opts_chunk$set(eval = have_beam, collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE,
                      dpi = 140, out.width = "100%", fig.width = 7, fig.height = 4.5)
```

[Three checks](https://imallona.github.io/beam/docs/explanations/metric-set-diagnostics.html) on a group of metrics used as one criterion, all based on the Spearman correlations between the metrics, here on the bundled OpenProblems batch integration scores. The scIB metrics are grouped into biological conservation and batch correction. The R functions call the Python `beam` package through reticulate; `install_beam_python()` installs it.

```{r libs}
library(rbeam)
datasets <- reticulate::import("beam.datasets")
cards <- reticulate::import("beam.cards")

op <- datasets$load_openproblems("batch_integration")
metrics <- setdiff(unlist(op$metric_ids), "hvg_overlap")
tensor <- op$tensor(metrics)
polarity <- unlist(cards$polarities_for(metrics))
bio <- c("ari", "nmi", "asw_label", "isolated_label_f1",
         "isolated_label_asw", "cell_cycle_conservation", "clisi")
groups <- ifelse(metrics %in% bio, "bio", "batch")
```

## All three checks

`beam_metric_diagnostics` runs validity, reliability and dimensionality on one grouping.

```{r diagnostics}
diag <- beam_metric_diagnostics(tensor, polarity, groups, metric_ids = metrics)
data.frame(
  group = names(diag$reliability$alpha_by_group),
  alpha = round(unlist(diag$reliability$alpha_by_group), 2),
  mean_inter_item = round(unlist(diag$reliability$mean_inter_item_by_group), 2),
  factors = unlist(diag$dimensionality$parallel_components_by_group),
  row.names = NULL
)
```

## Validity

Correlations within and between groups (Campbell and Fiske 1959):

```{r validity, fig.height = 5}
v <- diag$validity
cat(sprintf("mean within-group correlation:  %.2f\n", v$mean_convergent))
cat(sprintf("mean between-group correlation: %.2f\n", v$mean_discriminant))
cat("the split separates (within > between):", v$discriminant_ok, "\n")
beam_plot(v, "metric_correlation")
```

Metrics are ordered by group.

## Reliability

Standardized Cronbach's alpha per group (Cronbach 1951), with 0.7 as the usual cutoff.

```{r reliability, fig.height = 3.6}
rel <- diag$reliability
beam_plot(rel, "metric_reliability_dropped")
```

Each bar is the group alpha with one metric removed. The dashed line is the group alpha and the dotted line 0.7.

## Dimensionality

`metric_dimensionality` counts the factors of each group by parallel analysis (Horn 1965, Glorfeld 1995). The scree plot has the eigenvalues per group.

```{r scree, fig.height = 4}
dim_rep <- diag$dimensionality
beam_plot(dim_rep, "metric_dimensionality_scree")
```

The legend gives the number of factors retained by parallel analysis.

## Specification curve and blind analysis

The [specification curve](https://imallona.github.io/beam/docs/explanations/rank-sensitivity.html#the-specification-curve) over every weighting, aggregation and dataset. COMET is left out: it is slow with many metrics.

```{r spec-curve}
mcda <- reticulate::import("beam.mcda")
method_names <- unlist(op$method_names)
ctx <- mcda$registry_context(as.list(metrics), "saw")
rs <- beam_rank_sensitivity(
  tensor, ctx$polarity,
  methods = list("saw", "topsis", "vikor", "promethee_ii"),
  normalization = ctx$normalization, bounds = ctx$bounds,
  baselines = ctx$baselines, targets = ctx$targets,
  missing = "worst", tool_names = method_names, dataset_names = op$dataset_names
)
curve <- beam_specification_curve(rs)
top <- curve$tool_names[[curve$most_frequent_top_tool + 1]]
cat(sprintf("specification curve: %s ranks first in %.0f%% of %d combinations, %d reach the top\n",
            top, 100 * curve$most_frequent_top_fraction,
            curve$n_specifications, curve$n_distinct_top_tools))
```

```{r spec-curve-plot, fig.height = 5}
beam_plot(curve, "specification_curve")
```

A [blind analysis](https://imallona.github.io/beam/docs/explanations/analysis-blinding.html) fixes the weighting and the metric set before the method names are known. `beam_blind` masks the names; `beam_unblind` restores them.

```{r blind}
beam <- reticulate::import("beam")
scores <- beam$Scores(values = tensor,
                      tool_names = reticulate::tuple(as.list(method_names)),
                      metric_ids = reticulate::tuple(as.list(metrics)),
                      dataset_names = op$dataset_names, layout = "long")
bl <- beam_blind(scores, seed = 0L)
run <- beam_rank(bl$scores, weights = "equal", method = "topsis",
                 missing = "worst", sensitivity = FALSE, seed = 0L)
restored <- beam_unblind(run, bl$seal)
cat("top method after unblinding:", restored$top_tool, "\n")
cat("blinding fingerprint:",
    substr(run$manifest[["blinding"]][["seal_sha256"]], 1, 12), "\n")
```

## Dataset concordance

[`beam_dataset_concordance`](https://imallona.github.io/beam/docs/explanations/dataset-concordance-and-discrimination.html) ranks the methods within each dataset and compares every pair of per-dataset orderings with Kendall tau-b.

```{r concordance, fig.height = 5}
conc <- beam_dataset_concordance(
  tensor, ctx$polarity,
  normalization = ctx$normalization, bounds = ctx$bounds,
  baselines = ctx$baselines, targets = ctx$targets,
  missing = "worst", tool_names = method_names, dataset_names = op$dataset_names,
  metric_ids = metrics
)
cat(sprintf("mean agreement across datasets (Kendall tau-b): %.2f\n",
            conc$mean_pairwise_tau))
beam_plot(conc, "dataset_concordance")
```

## Summary

Alpha is 0.85 for the biological group, with two factors under parallel analysis, and 0.62 for the batch group, with one. See `?beam_metric_validity`, `?beam_metric_reliability`, `?beam_metric_dimensionality`.
