---
title: "1. Ranking methods and checking the ranking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Ranking methods and checking the ranking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_beam <- requireNamespace("reticulate", quietly = TRUE) &&
  reticulate::py_module_available("beam.datasets")
knitr::opts_chunk$set(eval = have_beam, collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE,
                      dpi = 140, out.width = "100%", fig.width = 7, fig.height = 4.5)
```

The Duo, Robinson and Soneson (2018) single-cell clustering benchmark bundled with the package: 14 methods, 12 datasets, and the metrics ARI (higher is better), runtime and Shannon entropy difference (lower is better). The R functions call the Python `beam` package through reticulate; `install_beam_python()` installs it. See the [aggregation methods](https://imallona.github.io/beam/docs/explanations/aggregation-methods.html).

```{r libs}
library(rbeam)
datasets <- reticulate::import("beam.datasets")
cards <- reticulate::import("beam.cards")
beam <- reticulate::import("beam")

duo <- datasets$load_duo2018()
method_names <- unlist(duo$method_names)
dataset_names <- unlist(duo$dataset_names)
metrics <- c("ari", "runtime", "shannon_entropy_diff")
polarity <- unlist(cards$polarities_for(metrics))
```

## Ranking

`beam_rank` takes the method by dataset by metric tensor as a `Scores` object, pools it across datasets by the rule on each card, ranks, and runs the sensitivity analysis.

```{r rank}
scores <- beam$Scores(
  values = duo$tensor(metrics),
  tool_names = reticulate::tuple(as.list(method_names)),
  metric_ids = reticulate::tuple(as.list(metrics)),
  dataset_names = reticulate::tuple(as.list(dataset_names)),
  layout = "long"
)
run <- beam_rank(scores, weights = "equal", method = "saw", seed = 0L)
cat("top-ranked under equal weights / SAW:", run$top_tool, "\n")
```

The [funky heatmap](https://imallona.github.io/beam/docs/explanations/funky-heatmaps-and-robustness.html) of the run:

```{r funky, fig.width = 8, fig.height = 7.5}
plot(run, metric_groups = c("accuracy", "cost", "accuracy"))
```

## SMAA

Fraction of sampled weightings with each method ranked first:

```{r smaa, fig.height = 4}
beam_plot(run, "smaa")
```

## Leave-one-dataset-out

Each dataset is dropped in turn and the rest are pooled and re-ranked. The stability of a method is the fraction of those runs with its rank unchanged.

```{r lodo, fig.height = 4}
lodo <- run$leave_one_dataset_out
cat("most influential dataset:",
    lodo$dataset_names[[lodo$most_influential_dataset + 1]],
    "(largest rank shift", lodo$max_rank_shift, ")\n")
beam_plot(run, "dataset_stability")
```

## Standalone checks

Each check is also a function on a tool by metric matrix. Here the benchmark is pooled to one value per metric.

```{r standalone}
pooled <- sapply(seq_along(metrics), function(j) {
  m <- duo$tensor(list(metrics[j]))[, , 1]
  if (metrics[j] == "runtime") exp(rowMeans(log(m), na.rm = TRUE)) else rowMeans(m, na.rm = TRUE)
})
agree <- beam_aggregation_agreement(pooled, polarity)
pert <- beam_smallest_weight_perturbation(pooled, polarity)
loo <- beam_leave_one_metric_out(pooled, polarity, metric_ids = metrics)
cat("five aggregations agree on the top tool:", agree$top_is_unanimous, "\n")
cat("mean pairwise Kendall agreement across aggregations:",
    round(agree$mean_pairwise_tau, 2), "\n")
cat("the top rank is fragile to a small weight change:", pert$top_rank_is_fragile, "\n")
```

Tools above the chance baseline, and tool pairs within the noise floor on every metric:

```{r reference-levels}
cat("tools that score above chance on no metric:",
    length(run$random_baseline$tools_never_beating), "\n")
cat("the top two tools are within the noise floor:",
    run$noise_floor$top_pair_indistinguishable, "\n")
```

## Card and data consistency

[`beam_card_data_consistency`](https://imallona.github.io/beam/docs/explanations/card-data-consistency.html) checks the raw scores against the range, chance baseline and noise floor on each card.

```{r card-consistency}
mcda <- reticulate::import("beam.mcda")
ctx <- mcda$registry_context(as.list(metrics), "saw")
audit <- beam_card_data_consistency(
  pooled, ctx$polarity, ctx$bounds,
  baselines = ctx$baselines, targets = ctx$targets, noise_floors = ctx$noise_floors,
  metric_ids = metrics
)
cat("pooled scores consistent with the cards:", audit$ok, "\n")
```

## Rank sensitivity

[`beam_rank_sensitivity`](https://imallona.github.io/beam/docs/explanations/rank-sensitivity.html) ranks the methods on each dataset under every weighting and aggregation and splits the rank variance of each method into the dataset, the weighting, the aggregation and their interactions.

```{r rank-sensitivity}
rs <- beam_rank_sensitivity(
  duo$tensor(metrics), ctx$polarity,
  normalization = ctx$normalization, bounds = ctx$bounds,
  baselines = ctx$baselines, targets = ctx$targets,
  missing = "worst", tool_names = method_names, dataset_names = dataset_names
)
cat(sprintf("dataset %.2f, weighting %.2f, aggregation %.2f, interaction %.2f\n",
            rs$dataset_share, rs$weighting_share, rs$aggregation_share, rs$interaction_share))
cat("most influential factor:", rs$most_influential_factor, "\n")
```

The same split per method, with each method's rank range:

```{r rank-sensitivity-by-tool, fig.height = 5}
beam_plot(rs, "rank_sensitivity_by_tool")
```

## Dataset concordance

[`dataset_concordance`](https://imallona.github.io/beam/docs/explanations/dataset-concordance-and-discrimination.html) ranks the methods within each dataset and compares every pair of per-dataset orderings with Kendall tau-b. The standalone function is `beam_dataset_concordance(tensor, polarity)`.

```{r concordance}
conc <- run$dataset_concordance
cat(sprintf("mean agreement across datasets (Kendall tau-b): %.2f\n",
            conc$mean_pairwise_tau))
cat("least typical dataset:",
    conc$dataset_names[[conc$most_idiosyncratic_dataset + 1]], "\n")
```

The agreement matrix, and the rank of each method per dataset minus its mean rank:

```{r concordance-plots, fig.height = 5}
beam_plot(run, "dataset_struggle")
```

## Specification curve

[`beam_specification_curve`](https://imallona.github.io/beam/docs/explanations/rank-sensitivity.html#the-specification-curve) takes the report of `beam_rank_sensitivity` and reports how often the top method is the same.

```{r specification-curve}
curve <- beam_specification_curve(rs)
top <- curve$tool_names[[curve$most_frequent_top_tool + 1]]
cat(sprintf("%d specifications; %s ranks first in %.0f%%, %d method(s) reach the top\n",
            curve$n_specifications, top, 100 * curve$most_frequent_top_fraction,
            curve$n_distinct_top_tools))
```

```{r specification-curve-plot, fig.height = 5}
beam_plot(curve, "specification_curve")
```

## Blind analysis

A [blind analysis](https://imallona.github.io/beam/docs/explanations/analysis-blinding.html) fixes the weighting, the aggregation and the metric set before the method names are known. `beam_blind` hides the names and shuffles the rows; `beam_unblind` restores them. The seal fingerprint is recorded in the manifest.

```{r blinding}
blinded <- beam_blind(scores, seed = 0L)
blind_run <- beam_rank(blinded$scores, weights = "equal", method = "saw",
                       sensitivity = FALSE, seed = 0L)
restored <- beam_unblind(blind_run, blinded$seal)
cat("top method after unblinding:", restored$top_tool, "\n")
cat("blinding fingerprint:",
    substr(blind_run$manifest[["blinding"]][["seal_sha256"]], 1, 12), "\n")
```

## Critical difference on ARI

[`beam_critical_difference`](https://imallona.github.io/beam/docs/explanations/critical-difference.html) runs the Friedman test and the Nemenyi post-hoc on ARI, for the methods observed on every dataset.

```{r cd, fig.width = 7, fig.height = 3.6}
ari <- duo$tensor(list("ari"))[, , 1]
complete <- !apply(is.na(ari), 1, any)
block <- ari[complete, ]
block_names <- method_names[complete]
cd <- beam_critical_difference(block, higher_is_better = TRUE, tool_names = block_names)
cat(sprintf("Friedman p = %.1e, Nemenyi CD = %.2f\n",
            cd$friedman_pvalue, cd$critical_difference))
beam_plot(cd, "critical_difference")
```

[`beam_skillings_mack`](https://imallona.github.io/beam/docs/explanations/critical-difference.html#skillings-mack-for-incomplete-blocks) is the Friedman-type test for the incomplete ARI matrix.

```{r skillings-mack}
sm <- beam_skillings_mack(ari, higher_is_better = TRUE, method_names = method_names)
cat(sprintf("Skillings-Mack chi-squared = %.2f on %d df, p = %.1e\n",
            sm$statistic, as.integer(sm$df), sm$p_value))
```

## Pairwise superiority on ARI

[`beam_pairwise_superiority`](https://imallona.github.io/beam/docs/explanations/pairwise-method-comparison.html) compares the methods two at a time across the datasets, with the ARI noise floor as the equivalence band. The probability of superiority is the fraction of datasets with one method scoring higher than the other.

```{r pairwise}
ari_floor <- cards$properties_for(list("ari"))[[1]]$noise_floor
sup <- beam_pairwise_superiority(ari, "higher_is_better",
                                 rope = if (is.null(ari_floor)) 0 else ari_floor,
                                 method_names = method_names)
cat("highest standing:", method_names[sup$order[1] + 1],
    sprintf("(%.2f)\n", sup$standing[sup$order[1] + 1]))
cat("pairs the sign test cannot separate:",
    length(sup$equivalent_pairs), "of", length(sup$per_pair), "\n")
```

## Pairwise transitivity

[`beam_pairwise_transitivity`](https://imallona.github.io/beam/docs/explanations/pairwise-method-comparison.html#transitivity) reports the method preferred to every other one, the circular triads, and whether the relation is transitive.

```{r transitivity}
trans <- beam_pairwise_transitivity(sup)
choice <- if (is.null(trans$condorcet_choice)) "none" else method_names[trans$condorcet_choice + 1]
cat("method preferred to every other one:", choice, "\n")
cat("transitive:", trans$is_transitive,
    "; circular triads:", trans$n_circular_triads, "of", trans$n_triads, "\n")
```

## Bayesian sign comparison

[`beam_bayesian_sign_comparison`](https://imallona.github.io/beam/docs/explanations/pairwise-method-comparison.html#bayesian-sign-comparison) gives, for each pair, the posterior probabilities that one method is practically better, that the two are equivalent within the floor, or that the other is practically better (Benavoli et al. 2017).

```{r bayesian, fig.height = 5}
bayes <- beam_bayesian_sign_comparison(sup)
decisive <- sum(vapply(bayes$per_pair, function(p) p$decision != "inconclusive", logical(1)))
cat("pairs with a decisive posterior at 0.95:", decisive, "of", length(bayes$per_pair), "\n")
beam_plot(bayes, "bayesian_comparison")
```

See `?beam_rank`, `?beam_smaa`, `?beam_critical_difference`.
