---
title: "1. Ranking methods and checking the ranking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Ranking methods and checking the ranking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
have_beam <- requireNamespace("reticulate", quietly = TRUE) &&
  reticulate::py_module_available("beam.datasets")
knitr::opts_chunk$set(eval = have_beam, collapse = TRUE, comment = "#>")
```

beam ranks methods from a tool by metric score matrix with multi-criteria decision analysis, then qualifies the ranking with sensitivity and statistical checks. This vignette runs the pipeline on the Duo, Robinson and Soneson (2018) single-cell clustering benchmark bundled with the package: 14 methods, 12 datasets, and the three well-populated metrics ARI (higher is better), runtime and Shannon entropy difference (lower is better). The R functions forward to the Python `beam` core through reticulate; install it once with `install_beam_python()`.

```{r libs}
library(rbeam)
datasets <- reticulate::import("beam.datasets")
cards <- reticulate::import("beam.cards")
beam <- reticulate::import("beam")

duo <- datasets$load_duo2018()
method_names <- unlist(duo$method_names)
dataset_names <- unlist(duo$dataset_names)
metrics <- c("ari", "runtime", "shannon_entropy_diff")
polarity <- unlist(cards$polarities_for(metrics))
```

## One ranking, end to end

`beam_rank` reads the method by dataset by metric tensor as a `Scores` object, pools it across datasets by each card's rule, runs the MCDA pipeline, and runs the default sensitivity analysis on the same normalization context.

```{r rank}
scores <- beam$Scores(
  values = duo$tensor(metrics),
  tool_names = reticulate::tuple(as.list(method_names)),
  metric_ids = reticulate::tuple(as.list(metrics)),
  dataset_names = reticulate::tuple(as.list(dataset_names)),
  layout = "long"
)
run <- beam_rank(scores, weights = "equal", method = "saw", seed = 0L)
cat("top-ranked under equal weights / SAW:", run$top_tool, "\n")
```

The funky heatmap shows the run as a glyph table: methods as rows, metrics as circles sized by the card-resolved normalized score, a composite bar, and the rank-robustness panels.

```{r funky, fig.width = 7, fig.height = 5}
fig_path <- tempfile(fileext = ".png")
plot(run, fig_path)
knitr::include_graphics(fig_path)
```

## SMAA: does the top method hold across weightings?

SMAA samples weight vectors and reports, per method, the share of weightings that rank it first.

```{r smaa, fig.width = 7, fig.height = 4}
conf <- run$smaa$confidence_factor
ord <- order(conf)
barplot(conf[ord] * 100, names.arg = method_names[ord], horiz = TRUE, las = 1,
        col = "seagreen", xlab = "share of random weightings ranked first (%)",
        main = "SMAA acceptability", xlim = c(0, 100), cex.names = 0.7)
```

## Leave-one-dataset-out: does the order depend on one dataset?

For each dataset dropped, the rest are pooled and re-ranked. The per-method stability is the share of those runs in which the method keeps its base rank.

```{r lodo, fig.width = 7, fig.height = 4}
lodo <- run$leave_one_dataset_out
base_rank <- run$result$ranks
ord <- order(base_rank)
barplot(lodo$rank_stability[ord] * 100, names.arg = method_names[ord], horiz = TRUE,
        las = 1, col = "darkorange", xlim = c(0, 100), cex.names = 0.7,
        xlab = "rank held across leave-one-dataset-out runs (%)",
        main = "Leave-one-dataset-out stability")
cat("most influential dataset:",
    lodo$dataset_names[[lodo$most_influential_dataset + 1]],
    "(largest rank shift", lodo$max_rank_shift, ")\n")
```

## The checks are callable on their own

Everything above runs through `beam_rank`, but each check is a standalone function on a tool by metric matrix. Here the benchmark is pooled to one value per metric, then read by the individual wrappers.

```{r standalone}
pooled <- sapply(seq_along(metrics), function(j) {
  m <- duo$tensor(list(metrics[j]))[, , 1]
  if (metrics[j] == "runtime") exp(rowMeans(log(m), na.rm = TRUE)) else rowMeans(m, na.rm = TRUE)
})
agree <- beam_aggregation_agreement(pooled, polarity)
pert <- beam_smallest_weight_perturbation(pooled, polarity)
loo <- beam_leave_one_metric_out(pooled, polarity, metric_ids = metrics)
cat("five aggregations agree on the top tool:", agree$top_is_unanimous, "\n")
cat("mean pairwise Kendall agreement across aggregations:",
    round(agree$mean_pairwise_tau, 2), "\n")
cat("the top rank is fragile to a small weight change:", pert$top_rank_is_fragile, "\n")
```

`beam_rank` also reads the raw scores against the reference levels on the cards: how many tools score above the chance baseline, and which tool pairs no metric separates above its noise floor.

```{r reference-levels}
cat("tools that score above chance on no metric:",
    length(run$random_baseline$tools_never_beating), "\n")
cat("the top two tools are within the noise floor:",
    run$noise_floor$top_pair_indistinguishable, "\n")
```

## Do the scores honour the cards?

`beam_card_data_consistency` reads the raw scores against what each card declares (range, chance baseline, noise floor), before normalization. A value outside its declared range, or a metric on the wrong scale, is named here rather than distorting the ranking silently. The normalization context comes from the registry.

```{r card-consistency}
mcda <- reticulate::import("beam.mcda")
ctx <- mcda$registry_context(as.list(metrics), "saw")
audit <- beam_card_data_consistency(
  pooled, ctx$polarity, ctx$bounds,
  baselines = ctx$baselines, targets = ctx$targets, noise_floors = ctx$noise_floors,
  metric_ids = metrics
)
cat("pooled scores consistent with the cards:", audit$ok, "\n")
```

## What moves the ranking: the dataset or the choice?

`beam_rank_sensitivity` splits the ranking's variance between the data and the analyst's choices. It ranks the methods on each dataset, under every weighting and aggregation, and attributes each method's rank variance to the dataset, the weighting, the aggregation and their interactions. The shares sum to one.

```{r rank-sensitivity}
rs <- beam_rank_sensitivity(
  duo$tensor(metrics), ctx$polarity,
  normalization = ctx$normalization, bounds = ctx$bounds,
  baselines = ctx$baselines, targets = ctx$targets,
  missing = "worst", tool_names = method_names, dataset_names = dataset_names
)
cat(sprintf("dataset %.2f, weighting %.2f, aggregation %.2f, interaction %.2f\n",
            rs$dataset_share, rs$weighting_share, rs$aggregation_share, rs$interaction_share))
cat("most influential factor:", rs$most_influential_factor, "\n")
```

The dataset carries the most, close to the dataset variance share the mixed-effects model reads in the heterogeneity vignette, by a different method. This is not in tension with the stable leave-one-dataset-out result above: that asks whether the pooled order survives dropping one dataset, and it does, because pooling averages over the per-dataset disagreement. The dataset share measures that disagreement directly.

The shares above are pooled over the methods. The `rank_sensitivity_by_tool` plot splits the same variance one method at a time. It separates a method whose rank depends on the dataset from one that depends on the weighting or the aggregation. The span next to each bar is the difference between the method's best and worst rank.

```{r rank-sensitivity-by-tool, fig.width = 7, fig.height = 4}
fig_path <- tempfile(fileext = ".png")
beam_plot(rs, "rank_sensitivity_by_tool", fig_path)
knitr::include_graphics(fig_path)
```

## Dataset concordance: do the datasets agree on the order?

`beam_rank` also reports how the datasets agree among themselves. `dataset_concordance` ranks the methods within each dataset and compares every pair of per-dataset orderings with Kendall tau-b. A high mean says the pooled ranking stands in for the individual datasets; a low mean says it does not. The standalone wrapper is `beam_dataset_concordance(tensor, polarity)`.

```{r concordance}
conc <- run$dataset_concordance
cat(sprintf("mean agreement across datasets (Kendall tau-b): %.2f\n",
            conc$mean_pairwise_tau))
cat("least typical dataset:",
    conc$dataset_names[[conc$most_idiosyncratic_dataset + 1]], "\n")
```

The agreement matrix and the rank-deviation map plot from the run. The deviation map marks where each method places higher or lower than its own typical rank, the method-by-dataset cells that drive the disagreement.

```{r concordance-plots, fig.width = 7, fig.height = 4}
fig_path <- tempfile(fileext = ".png")
beam_plot(run, "dataset_struggle", fig_path)
knitr::include_graphics(fig_path)
```

## Specification curve

`beam_specification_curve` reads the same grid as a list of rankings and reports how often the top method holds. It takes the report from `beam_rank_sensitivity`.

```{r specification-curve}
curve <- beam_specification_curve(rs)
top <- curve$tool_names[[curve$most_frequent_top_tool + 1]]
cat(sprintf("%d specifications; %s ranks first in %.0f%%, %d method(s) reach the top\n",
            curve$n_specifications, top, 100 * curve$most_frequent_top_fraction,
            curve$n_distinct_top_tools))
```

## Blind analysis

A blind analysis fixes the weighting, the aggregation and the metric set before the method names are revealed. `beam_blind` hides the names and shuffles the rows; `beam_unblind` restores them. The ranking does not change, and the seal fingerprint is recorded in the manifest.

```{r blinding}
blinded <- beam_blind(scores, seed = 0L)
blind_run <- beam_rank(blinded$scores, weights = "equal", method = "saw",
                       sensitivity = FALSE, seed = 0L)
restored <- beam_unblind(blind_run, blinded$seal)
cat("top method after unblinding:", restored$top_tool, "\n")
cat("blinding fingerprint:",
    substr(blind_run$manifest[["blinding"]][["seal_sha256"]], 1, 12), "\n")
```

## Are the methods separable? Critical difference

The composite gives one ranking but does not say whether the methods differ beyond chance. `beam_critical_difference` runs the Friedman test and the Nemenyi post-hoc on the ARI block of methods observed on every dataset. Two methods whose average ranks lie within the critical difference are not separable.

```{r cd, fig.width = 7, fig.height = 3.6}
ari <- duo$tensor(list("ari"))[, , 1]
complete <- !apply(is.na(ari), 1, any)
block <- ari[complete, ]
block_names <- method_names[complete]
cd <- beam_critical_difference(block, higher_is_better = TRUE, tool_names = block_names)

ord <- order(cd$average_ranks)
ranks_sorted <- cd$average_ranks[ord]
names_sorted <- block_names[ord]
plot(ranks_sorted, rep(0, length(ranks_sorted)), pch = 19, xlim = rev(range(cd$average_ranks)),
     ylim = c(-1, 0.6), axes = FALSE, xlab = "average rank on ARI (1 = best)", ylab = "",
     main = sprintf("Friedman p = %.1e, Nemenyi CD = %.2f",
                    cd$friedman_pvalue, cd$critical_difference))
axis(1)
text(ranks_sorted, 0.05, names_sorted, srt = 90, adj = 0, cex = 0.7)
cliques <- cd$cliques
for (k in seq_along(cliques)) {
  idx <- unlist(cliques[[k]]) + 1
  segments(min(cd$average_ranks[idx]), -0.1 * k, max(cd$average_ranks[idx]), -0.1 * k,
           col = "steelblue", lwd = 3)
}
```

When the matrix has gaps, `beam_skillings_mack` is the coverage-aware version of the same test, run on the full ARI matrix.

```{r skillings-mack}
sm <- beam_skillings_mack(ari, higher_is_better = TRUE, method_names = method_names)
cat(sprintf("Skillings-Mack chi-squared = %.2f on %d df, p = %.1e\n",
            sm$statistic, as.integer(sm$df), sm$p_value))
```

## How often does one method outperform another?

The critical-difference test says which methods are separable, not by how much or how often. `beam_pairwise_superiority` compares the methods two at a time across the datasets, with the ARI noise floor as the equivalence band: a method counts as outperforming another on a dataset only when the difference clears that floor. The probability of superiority is the fraction of datasets on which one method outperforms the other.

```{r pairwise}
ari_floor <- cards$properties_for(list("ari"))[[1]]$noise_floor
sup <- beam_pairwise_superiority(ari, "higher_is_better",
                                 rope = if (is.null(ari_floor)) 0 else ari_floor,
                                 method_names = method_names)
cat("highest standing:", method_names[sup$order[1] + 1],
    sprintf("(%.2f)\n", sup$standing[sup$order[1] + 1]))
cat("pairs the sign test cannot separate:",
    length(sup$equivalent_pairs), "of", length(sup$per_pair), "\n")
```

Most pairs are not distinguishable: across the twelve datasets the ARI differences are often inside the floor, so neither method outperforms the other decisively. This is the critical-difference reading as an effect size, and it agrees with it.

## Do the pairwise majorities agree with one order?

`beam_pairwise_transitivity` reads the same pairwise relation and asks whether one order is consistent with it: the method preferred to every other one, the circular triads, and whether the relation is transitive.

```{r transitivity}
trans <- beam_pairwise_transitivity(sup)
choice <- if (is.null(trans$condorcet_choice)) "none" else method_names[trans$condorcet_choice + 1]
cat("method preferred to every other one:", choice, "\n")
cat("transitive:", trans$is_transitive,
    "; circular triads:", trans$n_circular_triads, "of", trans$n_triads, "\n")
```

## How sure are we that one method is better?

The tests above report a p-value. `beam_bayesian_sign_comparison` reads the same pairwise relation on the probability scale. For each pair it gives the posterior probability that one method is practically better, that the two are practically equivalent within the floor, or that the other is practically better (Benavoli et al. 2017). Plot it with `beam_plot(bayes, "bayesian_comparison")`.

```{r bayesian}
bayes <- beam_bayesian_sign_comparison(sup)
decisive <- sum(vapply(bayes$per_pair, function(p) p$decision != "inconclusive", logical(1)))
cat("pairs with a decisive posterior at 0.95:", decisive, "of", length(bayes$per_pair), "\n")
beam_plot(bayes, "bayesian_comparison")
```

## Reading

Seurat ranks first under the default pipeline and holds the top in almost every random weighting and every leave-one-dataset-out run, so the recommendation does not depend on the weighting or on any single dataset. The critical-difference diagram qualifies this: on ARI alone most methods sit in overlapping cliques, so the upper-middle methods are not separable at twelve datasets. See `?beam_rank`, `?beam_smaa`, `?beam_critical_difference` for the function documentation.
